# Assignment 2 

This week I have worked with Data wrangling exercises as well as data analyses of linear regression. I have learnt basics of summarization and plotting data, as well as the linear regression analysis.

The work of the week are presented next.

```{r}
date()

#2.0
#data wrangling in R script:

#Marianne Holopainen, 7.11.2022, exercise 2. SEE THESE IN create_learning2014.R

#reading the JYTOPKYS3.data into R.

#library(tidyverse) # install.packages("tidyverse")
#JYTOPKYS3_data <- read.delim("~/IODS-project/data/JYTOPKYS3-data.txt")
  #stringsAsFactors = FALSE
  #View(JYTOPKYS3_data)

#Create an analysis dataset with the variables gender, age, attitude, 
#deep, stra, surf and points.
#Scale all combination variables to the original scales (by taking the mean). 
#Exclude observations where the exam points variable is zero. 
#(The data should then have 166 observations and 7 variables)

#Deep     Deep approach             ~d_sm+d_ri+d_ue
#Surf     Surface approach          ~su_lp+su_um+su_sb
#Stra     Strategic approach        ~st_os+st_tm
#Points   Yhteispisteet (max kaikista)'

#VAR d_sm:1=D03+D11+D19+D27 #seaking meaning.
#+VAR d_ri:1=D07+D14+D22+D30 #relating ideas.
#+VAR d_ue:1=D06+D15+D23+D31 #use of evidence.
#+VAR su_lp:1=SU02+SU10+SU18+SU26 #lack of purpose.
#+VAR su_um:1=SU05+SU13+SU21+SU29 #unrelated memorising.
#+VAR su_sb:1=SU08+SU16+SU24+SU32 #syllabus-boundeness.
#+VAR st_os:1=ST01+ST09+ST17+ST25 #organized studying.
#+VAR st_tm:1=ST04+ST12+ST20+ST28 #time management.
#+VAR Deep:1=d_sm+d_ri+d_ue #deep approach.
#+VAR Surf:1=su_lp+su_um+su_sb #surface approach.
#+VAR Stra:1=st_os+st_tm #strategic approach.

#deep, surf, stra and their means. 
#attach(JYTOPKYS3_data)
#  JYTOPKYS3_data$deep <- (D03+D11+D19+D27+D07+D14+D22+D30+D06+D15+D23+D31)/12
#JYTOPKYS3_data$surf <- (SU02+SU10+SU18+SU26+SU05+SU13+SU21+SU29+SU08+SU16+SU24+SU32)/12
#JYTOPKYS3_data$stra <- (ST01+ST09+ST17+ST25+ST04+ST12+ST20+ST28)/8
#detach(JYTOPKYS3_data)

#exercise_data2 <- JYTOPKYS3_data[c(57:63)]#choose the correct outcomes.

#attach(exercise_data2)#lopullinen data tehtavaan.
#exercise_data_final <- exercise_data2[ which(Points > 0),]#exludes 0 values.
#detach(exercise_data2)

#View(exercise_data_final)

#Use `str()` and `head()` to make sure that the structure of the data is correct?
#write_csv(exercise_data_final, "learning2014.csv")
#read_csv("learning2014.csv")

```


## 2.1. The structure and the dimensions of the data

Data set includes seven variables: gender, age, attitude, deep, stra, surf and points. There are 166 participants (110 female and 56 male) in the data. Age range was from 21 to 55 years. Over half of the participants were from 20 to 23 years old (N=88). 

The outcomes of deep, stra and surf depict deep, surface, and strategic approach of study skills from the measure of Approaches and Study Skills Inventory for Students. Deep-outcome is composed by summing 12 questions, that ask about "seeking meaning", "relating ideas", and "use of evidence", and then divided by the number of the items of the scale (range: 1-5). Surf-outcome is composed by summing 12 questions, that ask about "lack of purpose", "unrelated memorising", and "syllabus-boundeness", and then divided by the number of the items of the scale (range: 1-5). In addition, Stra-outcome includes the mean of 8 items, that measure "organized studying" and "time management". Lastly, there are Points-outcome that includes the total score from all the questions.


```{r}
#1, explore the sturucture and dimensions of the data.
library(tidyverse) # install.packages("tidyverse")
learning2014.data <- read.csv("~/IODS-project/data/learning2014-data.txt")
stringsAsFactors = FALSE
View(learning2014.data)

#N = gender
learning2014.data %>%
  count(gender, sort = TRUE)
#N =age
learning2014.data %>%
  count(age, sort = TRUE)


library(tidyverse)
library(gapminder) # dataset
library(finalfit)
library(broom)
library(ggplot2)

#statistics.
glimpse(learning2014.data)
ff_glimpse(learning2014.data) 


```



## 2.2.  A graphical overview of the data and summaries of the variables in the data

Summary of statistics for the outcomes are presented in Table following:

```{r}
ff_glimpse(learning2014.data)
```

Means by gender are presented here: 

```{r}


#create data frame for Means
df <- data.frame(gender = c('female', 'male'),
                 Points = c(22.3, 23.5),
                 Stra = c(3.20, 2.96),
                 Deep = c(3.66, 3.72),
                 Surf = c(2.83, 2.70))

#view data frame
df
```

Females have lower total scores than males. On average, females used also more strategic and surface approach of study skills and male used more deep approach.

Boxplots present the means by gender here:


```{r}
boxplot(learning2014.data$points ~ learning2014.data$gender,
xlab = "Gender",
ylab = "Points")
boxplot(learning2014.data$deep ~ learning2014.data$gender,
xlab = "Gender",
ylab = "Deep Approach")
boxplot(learning2014.data$stra ~ learning2014.data$gender,
xlab = "Gender",
ylab = "Strategic Approach")
boxplot(learning2014.data$surf ~ learning2014.data$gender,
xlab = "Gender",
ylab = "Surface Approach")
```


and histograms for the outcomes are displayed here for the total sample: 

```{r}
hist(learning2014.data$points,
breaks = 20, 
xlab = " Points",
main = " Histogram of total Points")

hist(learning2014.data$deep,
breaks = 20, 
xlab = " Points",
main = " Histogram of Deep Approach")

hist(learning2014.data$surf,
breaks = 20, 
xlab = " Points",
main = " Histogram of total Surface Approach")

hist(learning2014.data$stra,
breaks = 20, 
xlab = " Points",
main = " Histogram of Strategic Approach")
```

From visual inspections, the distributions of the outcomes are near to normal distribution, except for total Points that are distributed more evenly.


The relationships between the outcomes of Surf, Stra and Deep, and Poits are presented in the following plots:

```{r}
plot(learning2014.data$deep, learning2014.data$points,
xlab = "Deep Approach",
ylab = "Points")
plot(learning2014.data$stra, learning2014.data$points,
xlab = "Strategic Approach",
ylab = "Points")
plot(learning2014.data$surf, learning2014.data$points,
xlab = "Surface Approach",
ylab = "Points")
```

and here with the lines:

```{r}
ggplot(learning2014.data, aes(x = deep, y = points)) +
  geom_point() +
  stat_smooth()
ggplot(learning2014.data, aes(x = surf, y = points)) +
  geom_point() +
  stat_smooth()
ggplot(learning2014.data, aes(x = stra, y = points)) +
  geom_point() +
  stat_smooth()
``` 

Although the relationship between points and 3 outcomes is a bit less clear, it still appears somewhat linear. Linear regression can be proceed. 


## 2.3 and 2.4. Regression analyses.

Model 1:

```{r}
#First the correlations are checked:

#check the correlations
#cor(learning2014.data$stra, learning2014.data$deep)
#cor(learning2014.data$stra, learning2014.data$surf)
#cor(learning2014.data$surf, learning2014.data$deep)
```

Linear regression model 1 with three variables (deep, stra, surf) as explanatory variables and exam points as the target (dependent, outcome) variable are summarized below:

```{r}
model1 <- lm(points ~ deep + stra + surf, data = learning2014.data)

summary(model1)

``` 

The results show that the model is not statistically significant (F(3, 162) = 2.292, p = 0.08). The estimated effect of deep approach is -0.74 (p = 0.39), while the estimated effect for strategic approach is 0.99 (p = 0.10) and for surface approach -1.63 (0.08). 

In the model1 2.3 % of the varince in the points can be predicted by the three outcomes.


Model2:

In the model 2, deep approach outcome is excluded from the model 1, and it is summarized here: 

```{r}
model2 <- lm(points ~ stra + surf, data = learning2014.data)

summary(model2)

``` 

The results depict that the model is now statistically significant (F(2, 163) = 3.074, p = 0.049). The estimated effect for strategic approach is 0.96 (p = 0.11) and for surface approach -1.38 (p = 0.11). This means that more strategic approach and less surface approach predicts the higher the points. However, even though the model is statistically significant, the the estimated effect are insignificant for independent variables.*

In the model2, 2.5 % of the variance in the points (adjusted R-squared) can be predicted by the two outcomes.


## 2.5. The diagnostic plots.

```{r}
par(mfrow=c(2,2))
plot(model2)
par(mfrow=c(1,1))

```

There is the assumption of homoscedasticity in the model. 

In a plot of residual vs fitted values red line represent the mean of the residuals. It supposed to be horizontal and centered around zero. This means there are no outliers or biases in the data that would make a linear regression invalid.In the model2, there are a little bias in small and large values of the points.

In the Normal Q-Qplot in the top right, you can see that the real residuals from model2 form an almost perfectly one-to-one line with the theoretical residuals from a perfect model when theoretical qouantiles are [-1, 1]

Based on these residuals, we can say that our model meets the assumption of homoscedasticity at least to a certain extent.

In the plot residuals vs leverage is shown several influential points that are below Cook's distance. Leverage refers to the extent to which the coefficients in the regression model would change if a particular observation was removed from the dataset. Based on the results, there might be an option that another regression model would better fit in or the influential observations should be handle in another way.




