# Assignment 6

```{r}
#read the data sets from folder

library(dplyr)
library(tidyr)
library(tidyverse)

RATSL <- read.csv("~/IODS-project/RATSL.csv")
stringsAsFactors = FALSE
RATS <- read.csv("~/IODS-project/RATS.csv")
stringsAsFactors = FALSE
BPRSL <- read.csv("~/IODS-project/BPRSL.csv")
stringsAsFactors = FALSE
BPRS <- read.csv("~/IODS-project/BPRS.csv")
stringsAsFactors = FALSE


#Convert the categorical variables of both data sets to factors (again)
BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)



RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

#names
names(BPRS)
names(RATS)
names(BPRSL)
names(RATSL)

#structure
str(BPRS)
str(RATS)
str(BPRSL)
str(RATSL)

#summary
summary(BPRS)
summary(RATS)
summary(BPRSL)
summary(RATSL)

View(RATS)
View(BPRS)
View(RATSL)
View(BPRSL)

```


## 6.1 Analyses for RATS data

```{r}
#Graphical displays of longitudinal data

# Access the packages
library(dplyr)
library(tidyr)
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
   scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))


#data standardization and plots again for RATS data
# Standardise the variable Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised RATS

ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Weight")


# Summary data with mean and standard error of Weight by Group and Time 
n1 <- 16

RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight) / sqrt(n1) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)
View(RATSS)

# Plot the mean profiles
library(ggplot2)
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=2) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.5, color = 'red') +
  theme(legend.position = c(0.8,0.4)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")


#Outliners

#Draw the boxplot and observe the outlier
#Find a suitable threshold value and use `filter()` to exclude the outlier to form a new data
library(dplyr)
library(tidyr)

# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline)
RATSL_out <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL_out)

# Draw a boxplot of the mean versus treatment

ggplot(RATSL_out, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Times 8-64")

View(RATSL_out)

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL_out2 <- RATSL_out %>%
 filter(mean < 550) %>%
  filter (mean > 250)

#boxplot without outliner
ggplot(RATSL_out2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Times 8-64")


#t-tests for the filtered data:

df12 <- RATSL_out2 %>% 
  filter(Group == 1 | Group == 2)
t.test(mean ~ Group, data = df12, var.equal = TRUE)

df13 <- RATSL_out2 %>% 
  filter(Group == 1 | Group == 3)
t.test(mean ~ Group, data = df13, var.equal = TRUE)

df23 <- RATSL_out2 %>% 
  filter(Group == 3 | Group == 2)
t.test(mean ~ Group, data = df23, var.equal = TRUE)

#Results from an Analysis of Covariance of the RATS Data with Baseline
#RATS and Group as Covariates

RATSL_out2 <- RATSL_out %>%
  mutate(baseline = RATS$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL_out2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

-The above analyses initially explores the longitudinal RATS data. The data is from a nutrition study conducted in three groups of rats (N = 16). The three groups of RATS were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately weekly) over a 9-week period.

-The first figure shows a weight profiles by nutrition group for the RATS data, and the second plot displays the profiles after standardization. The first graph show that weight of rats is slightly increasing over the 9 week study period and rats to have higher weight at the beginning tend to have higher values throughout the study, approximately. Both graphs shows also that there are differences in baseline weight of rats by groups, especially in group1 rats seem to be lighter than in other groups. The standardized plot also seems to show that variability appears to be same over time. In addition, it indicates that there are differences between the two groups on means of weight but at the same time suggesting a little differences in a change of weight by time. 

-The mean response profiles for the three groups with error bars illustrates that group 1 and group 2 has only little variation between the rats in the group. The profiles also shows that in group 1 the rats are the lightest during the sutdy and in group 3 the heaviest. Also by time, there seems to be only a little change of mean weight.

-In the out-liner analysis I concluded to filter out the heaviest and the lightest rat.

-The two-sample t-tests confirms the differences of means between group 1 and  group 2, and group 1 and group3, but not between group 2 and 3.

-Results from an Analysis of Covariance  with baseline weight and nutrition Group as covariates show that baseline weight is strongly related to the weight values but no evidence of a nutrition treatment difference are not found even the baseline weight is conditioning. 




## 6.2 Analyses for BPRS data


```{r}



# Check the dimensions of the data
dim(BPRS)
View(BPRSL)

# Plot the BPRSL data
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))


# create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)


# access library lme4
library(lme4)

# Create a random intercept model (subject as random effect)
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)


# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment  + (week | subject) + week*treatment, data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

# draw the plot of BPRSL with the observed bprs values
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "bprs", limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))



# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

library(dplyr)
library(tidyr)
# Create a new column fitted 
BPRSL <- BPRSL %>% mutate(fitted = Fitted)

# draw the plot of RATSL with the Fitted values of weight
library(ggplot2)
ggplot(BPRSL, aes(x = week, y = fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(name = "Fitted bprs", limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
 
```

The BPRS data included 20 subjects that were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (wekk 0) and then at weekly intervals for eight weeks.The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having
schizophrenia.


To begin there is the plot for the BPRS values for all participants, differentiating between the treatment groups into which they have been randomized.

Next, there are presented results from fitting a linear regression model to BPRS Data with bprs as
response variable, and treatment group and week as explanatory variables (ignoring the Repeated-Measures Structure of the Data). Summary shows that week has significant effect but treatment does not. 


The second model take account for the independence of the repeated measures. In anova analysis is shown that there is a significant difference between the two models.


The last model allows for a treatment - week interaction to BPRS data. The likelihood ratio test of
the interaction random intercept and slope model against the corresponding model without an interaction is 3.17 with 1 DF; the associated p-value is >0.05, so it could be  concluded that the interaction model does not provide a better fit for the data.

Lastly, the graphics show how well the interaction model fits the observed data. In observed data, there are much more individual variation that fitted data. 













