# Assignment 5


## 5.1 Data description 

```{r}
library(tidyr)
library(stringr)


#read the datasets
human5 <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/human2.txt", sep=",", header = TRUE)

#colnames
colnames(human5)

#structure
str(human5)

#summary
summary(human5)


#distributions of variables and graphical overview
# visualize the dataset 
library(ggplot2)
pivot_longer(human5, cols = everything()) %>% 
  ggplot(aes(value)) + facet_wrap("name", scales = "free") + geom_histogram(bins = 30)


#relationships between variables
library(corrplot)
# compute the correlation matrix and visualize it with corrplot
cor(human5) %>% corrplot()
cor(human5)


```

The data includes 8 variables (and 155 observations):
  -Ado.Birth; Adolescent birth rate
  -Edu.Exp; Expected years of schooling
  -Edu2.FM; Population with Secondary Education (Female)/Population with Secondary Education (Male)
  -GNI; Gross National Income per capital
  -Labo.FM; Labour Force Participation Rate (Female)/Labour Force Participation Rate (Male)
  -Life.Exp; Life expectancy at birth
  -Mat.Mor; Maternal mortality ratio
  -Parli.F; Percetange of representatives in parliament

Mean of the life expectancy was  72 years and expected year of schooling 13 years.

There were on average more males in population with secondary education and in labour force.

In average, there was 21 representatives in parliament.

Life expectancy, Expected years of schooling, Maternal mortality ratio (negative) and Adolescent birth rate (negative) were highly correlated.


## 5.2 Principal component analysis (PCA)


```{r}
#RAW data
# perform principal component analysis (with the SVD method) on the raw human data
pca_human5 <- prcomp(human5)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human5, choices = 1:2, col = c("grey40", "deeppink2"), cex = c(0.5, 1))


#standardized data
# standardize the variables
human_std <- scale(human5)

# print out summaries of the standardized variables
summary(human_std)

# perform principal component analysis (with the SVD method)
pca_human5_std <- prcomp(human_std)

# draw a biplot of the principal component representation and the original variables
biplot(pca_human5_std, choices = 1:2, col = c("grey40", "deeppink2"), cex = c(0.4, 1))


```


Principal component analyses are presented above for raw data and standardized data. Results differed from each others, due to different scales used in outcomes. There were a lot of zero-length arrows in the raw-model, except GNI. 

My personal interpretations of the first two principal component dimensions based on the biplot drawn for PCA on the standardized human data are Health and Equality.


## 5.3 Tea dataset


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
tea <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/Helsinki-Open-Data-Science/master/datasets/tea.csv", sep = ",", header = T)

#structure of the data
colnames(tea)

View(tea)

str(tea)

#convert outcomes to factors
tea$breakfast <- factor(tea$breakfast)
tea$tea.time <- factor(tea$tea.time)
tea$evening <- factor(tea$evening)
tea$lunch <- factor(tea$lunch)
tea$dinner <- factor(tea$dinner)
tea$always <- factor(tea$always)
tea$home <- factor(tea$home)
tea$work <- factor(tea$work)
tea$tearoom <- factor(tea$tearoom)
tea$friends <- factor(tea$friends)
tea$resto <- factor(tea$resto)
tea$pub <- factor(tea$pub)
tea$Tea <- factor(tea$Tea)
tea$How <- factor(tea$How)
tea$sugar <- factor(tea$sugar)
tea$how <- factor(tea$how)
tea$sugar <- factor(tea$sugar)
tea$where <- factor(tea$where)
tea$Tea <- factor(tea$Tea)
tea$price <- factor(tea$price)
tea$sex <- factor(tea$sex)
tea$SPC <- factor(tea$SPC)
tea$Sport <- factor(tea$Sport)
tea$age_Q <- factor(tea$age_Q)
tea$frequency <- factor(tea$frequency)
tea$escape.exoticism <- factor(tea$escape.exoticism)
tea$spirituality <- factor(tea$spirituality)
tea$healthy <- factor(tea$healthy)
tea$diuretic <- factor(tea$diuretic)
tea$friendliness <- factor(tea$friendliness)
tea$iron.absorption <- factor(tea$iron.absorption)
tea$feminine <- factor(tea$feminine)
tea$sophisticated <- factor(tea$sophisticated)
tea$slimming <- factor(tea$slimming)
tea$exciting <- factor(tea$exciting)
tea$relaxing <- factor(tea$relaxing)
tea$effect.on.health <- factor(tea$effect.on.health)


library(dplyr)
library(tidyr)

colnames(tea)

# column names to keep in the dataset
keep_columns <- c("tea.time","evening", "lunch","dinner","always","home","work", "tearoom","friends","resto","pub","Tea",
"How","sugar","how","where","price","sex","SPC","Sport","age_Q" ,"frequency" 
,"escape.exoticism","spirituality","healthy","diuretic","friendliness","iron.absorption" 
,"feminine","sophisticated","slimming","exciting","relaxing","effect.on.health")

# select the 'keep_columns' to create a new dataset
tea_time <- dplyr::select(tea, keep_columns)

# look at the summaries and structure of the data
summary(tea_time)
str(tea_time)

# visualize the dataset 
library(ggplot2)
pivot_longer(tea_time, cols = everything()) %>% 
  ggplot(aes(value), theme(axis.text.x = element_text(angle = 120, hjust = 1, size = 3))) + facet_wrap("name", scales = "free") + geom_bar()

#multiple correspondence analysis
library(FactoMineR)
mca <- MCA(tea_time, graph = FALSE)

# summary of the model
summary(mca)

# visualize MCA
plot(mca, invisible=c("ind"), graph.type = "classic", habillage = "quali", cex = c(1.5, 0.2))

#multiple correspondence analysis for the selcected outcomes
library(FactoMineR)
mca2 <- MCA(tea_time[1:13], graph = FALSE)

# summary of the model
summary(mca2)

# visualize MCA
plot(mca2, invisible=c("ind"), graph.type = "classic", habillage = "quali")

install.packages("ggplot2")
library(ggplot2)
library(factoextra)




#to visualize the percentages of inertia explained by each MCA dimensions,
fviz_screeplot(mca, addlabels = TRUE, ylim = c(0, 45))
fviz_screeplot(mca2, addlabels = TRUE, ylim = c(0, 45))



```


The Multiple Correspondence Analysis (MCA) on the total tea data (mca) and for the selected variables are presented above as well as biplots and plots to visualize the percentages of inertia explained by each MCA dimensions.

In a mca2 model the first dimension explained 13.3 % of the inertia that seems to considered to  where and when tea is drunk. 




